---
title: "Cox Proportional Hazards Model"
author: "Halley Deleeuw and Jasmine Sawh"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction

This is an introduction to Cox Proportional Hazards regression, which is a statistical method used for survival analysis. It uses a semi-parametric estimator that focuses on estimating the hazard function which represents the instantaneous risk of an event occurring at a given time. It's mostly used in industries such as medical, epidemiology and various other fields where they need to analyze data pertaining to a failure or an event such as death or relapse. [Gonzalez-Dominguez] The purpose of Cox Proportional Hazards regression is to find the relationship between one or more independent variables and the hazard function while allowing for the exploration in non-linearity and time-varying effects. [W.W.M Abeysekera] The dependent variable in a hazard model consist of two parts, the first part is the event indicator and the second part is a measure of time. [Jane E.Miller] The purpose of this paper is to take a dive deeper into this regression model by analyzing the (insert data set here). This data set will use the regression to [add pertaining information here]. 

(This is a note for myself so I know where I got each information from so I can properly cite it later.)
Gonzalez-Dominguez Cox Proportional Hazards model used for predictive analysis of the energy consumption of healthcare buildings. 
W.W.M Abeysekera Use of Schoenfeld’s global test to test the proportional hazards
assumption in the Cox proportional hazards model: an application
to a clinical study


## Methods

The common non-parametric regression model is
$Y_i = m(X_i) + \varepsilon_i$, where $Y_i$ can be defined as the sum of
the regression function value $m(x)$ for $X_i$. Here $m(x)$ is unknown
and $\varepsilon_i$ some errors. With the help of this definition, we
can create the estimation for local averaging i.e. $m(x)$ can be
estimated with the product of $Y_i$ average and $X_i$ is near to $x$. In
other words, this means that we are discovering the line through the
data points with the help of surrounding data points. The estimation
formula is printed below [@R-base]:

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$ $W_n(x)$ is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if $X_i$ is far from $x$.

## Analysis and Results

### Data and Vizualisation

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Statistical Modeling

### Conlusion

## References
